---
title: "MDR_RR_TB_burden_analysis"
author: "Lun Suša"
format: html
editor: visual
---

```{r}
library(tidyverse)
library(dplyr)

```

```{r}
mdr_rr_data <- read.csv("../_raw/MDR_RR_TB_burden_estimates.csv")

TB_dictionary <- read.csv("../_raw/TB_data_dictionary.csv")


```

**Cleaning of MDR_RR_TB_burden:**

Get the descriptions of the variables from the dictionary to understand what we are looking at:

```{r}
df_colnames <- tibble(colname = colnames(mdr_rr_data))

dict_filtered <- TB_dictionary[c("variable_name", "definition")] |> 
   filter(variable_name %in% (df_colnames |> 
                                pull(colname)))
dict_filtered

dict_missing <- df_colnames |> 
  filter(!colname %in% (dict_filtered |>  
           pull(variable_name)))
dict_missing

```

The dictionary covers all variables except year (which should be obvious to me but was not).\
\

Renaming them to be more intuitive:\

```{r}
mdr_rr_data <- mdr_rr_data |> 
  rename(
    
    # Estimated RR-TB incidence (absolute)
    rr_incidence = e_inc_rr_num,
    rr_incidence_max = e_inc_rr_num_hi,
    rr_incidence_min = e_inc_rr_num_lo,
    
    # Estimated RR-TB among lab-confirmed pulmonary TB
    rr_labconf_pulm = e_rr_in_notified_labconf_pulm,
    rr_labconf_pulm_max = e_rr_in_notified_labconf_pulm_hi,
    rr_labconf_pulm_min = e_rr_in_notified_labconf_pulm_lo,
    
    # % of new TB cases
    rr_new = e_rr_pct_new,
    rr_new_max = e_rr_pct_new_hi,
    rr_new_min = e_rr_pct_new_lo,
    
    # % of previously treated cases with RR-TB
    rr_treated = e_rr_pct_ret,
    rr_treated_max = e_rr_pct_ret_hi,
    rr_treated_min = e_rr_pct_ret_lo,
    
    #method for new TB
    rr_method_new = source_rr_new,
    
    #method for previously treated TB
    rr_method_treat = source_rr_ret
  )
```

What is the number of observations per year and per country?\

```{r}
mdr_rr_data |> count(country)
mdr_rr_data |> count(year)
```

So we have 215 countries ans 10 years, with 1 observation per year per country.\
\

Are there missing values?\

```{r}
mdr_rr_data |> 
  summarise(across(everything(),
                   ~sum(is.na(.))))

```

All estimated values have 330 or 345 missing values.\

```{r}
mdr_rr_data |> filter(!is.na(iso2)) |> 
               filter(if_any(everything(), is.na)) |> 
                slice_sample(n=20)

mdr_rr_data |> filter(is.na(iso2)) 
```

Random observations are missing estimated values.\
Namibia is missing 'iso2'.\

Removing 'iso2' and other reduntant variables as with other datasets:

```{r}
mdr_rr_data <- mdr_rr_data |> 
  select(-iso2, -iso3, -iso_numeric)
```

Are there countries that are entirely missing estimated values for all observations?\

```{r}
mdr_rr_data|> 
  group_by(country) |> 
  summarise(all_missing = all(is.na(rr_incidence))) |> 
  filter(all_missing)
```

33 out of 215 countries are missing all estimated variables. Not going to do anything about it for now, but worth noting.
