---
title: "04_describe"
format: html
---

WHO 2025 data on Tuberculosis, data description

In this script, it's intended to give an explanation on data features, using different methods and graphs. The provided data description is based on three datasets:

-   The joined dataset which only haves information from 2024
-   The joined dataset which contains information from 2015 to 2024
-   The WHO TB burden estimates \[\>1Mb\] dataset, as it contains information from previous years

------------------------------------------------------------------------

**Note to self - add this in all code chunks when wrapping up:**

```{r}
#| echo: true
#| eval: true
```

**Loading relevant libraries:**

```{r}
library("tidyverse") 
```

```{r}
#Access to the function for loading the datasets and to save them
source("99_proj_func.R")
```

------------------------------------------------------------------------

**Loading data:**

Joined dataset, year 2024:

```{r}
#Loading age and sex and risk group data (2024 only): 

data_file          <- "TB_age_sex_joined.tsv"
TB_age_sex_joined  <- load_data(data_file)

#Display the data: 
TB_age_sex_joined
```

Joined dataset, year 2015 to 2024:

```{r}
#Loading age and sex and risk group data (2024 only): 

data_file           <- "TB_10_years_joined.tsv"
TB_10_years_joined  <- load_data(data_file)

#Display the data: 
TB_10_years_joined
```

WHO TB burden estimates \[\>1Mb\]:

```{r}
#Loading TB burden estimates: 

data_file   <- "TB_burden.tsv"
TB_burden   <- load_data(data_file)

#Display the data: 
TB_burden
```

------------------------------------------------------------------------

It's late, hope someone sees this before me, maybe some statistics and graphs would be fine, but nothing that complex, complex stuff should go into analysis part.

**Description for TB_age_sex_joined**

This dataset contains the number of TB cases across different countries, categorized by age group and gender. It also includes cases per 100,000 population, enabling standardized and comparable analysis between countries.

TB cases per 100k

```{r}
# Filter the data
TB_top <- TB_age_sex |> 
  filter(country %in% top_10_countries)

# Countries with the top 10 most TB cases pr. 100k:
ggplot(TB_top, aes(y = country, x = TB_cases_best)) +
  geom_boxplot(alpha = 0.7, fill = "orange") +
  labs(x = "TB cases\n(Separated based on different metadata rows)", y = "Country") +
  theme_minimal()
```
