---
title: "05_analysis_1"
format:
  html:
    self-contained: true
---

**Loading relevant libraries:**

```{r}
library(tidyverse) 
library(dplyr)
library(ggplot2)
library(maps)
library(scales)

#Access to the function for loading the datasets and to save them
source("99_proj_func.R")
```

------------------------------------------------------------------------

# **Loading data:**

**Joined dataset, year 2015 to 2024:**

```{r}
#Loading age and sex and risk group data (2024 only): 

data_file           <- "03_aug_TB_10_years.tsv"
TB_10_years_joined  <- load_data(data_file)

#Display the data: 
TB_10_years_joined
```

**Dictionary:**

```{r}
#Loading the data dictionary: 

data_file   <- "01_load__ictionary.tsv"
TB_dictionary <- load_data(data_file)
TB_dictionary

```

**Mutating the country names:**

In this analysis we will plot the data on a world map using "library(maps)", hence we need to change the country names to match the names in the package. We create a map dataset for all the years.

```{r}
# 1. Load the world map coordinates
world_map <- map_data("world")

# 2. Create the main map dataframe
map_TB_10_years_joined <- world_map |> 
  left_join(
    TB_10_years_joined |>
      # Create a 'region' column that matches the map library names
      mutate(region = case_when(
        country == "United States of America"   ~ "USA",
        country == "Bolivia (Plurinational State of)" ~ "Bolivia",
        country == "Brunei Darussalam"          ~ "Brunei",
        country == "Cabo Verde"                 ~ "Cape Verde",
        country == "China, Hong Kong SAR"       ~ "Hong Kong",
        country == "China, Macao SAR"           ~ "Macao", 
        country == "Congo"                      ~ "Republic of Congo",
        country == "Côte d'Ivoire"              ~ "Ivory Coast",
        country == "Czechia"                    ~ "Czech Republic",
        country == "Democratic People's Republic of Korea" ~ "North Korea",
        country == "Eswatini"                   ~ "Swaziland",
        country == "Iran (Islamic Republic of)" ~ "Iran",
        country == "Lao People's Democratic Republic" ~ "Laos",
        country == "Netherlands (Kingdom of the)" ~ "Netherlands",
        country == "occupied Palestinian territory, including east Jerusalem" ~ "Palestine",
        country == "Republic of Korea"          ~ "South Korea",
        country == "Republic of Moldova"        ~ "Moldova", 
        country == "Russian Federation"         ~ "Russia",
        country == "Sint Maarten (Dutch part)"  ~ "Sint Maarten",
        country == "Syrian Arab Republic"       ~ "Syria",
        country == "Tokelau"                    ~ "Tokelau",
        country == "Trinidad and Tobago"        ~ "Trinidad",
        country == "Türkiye"                    ~ "Turkey",
        country == "Tuvalu"                     ~ "Tuvalu",
        country == "United Kingdom of Great Britain and Northern Ireland" ~ "UK",
        country == "United Republic of Tanzania" ~ "Tanzania",
        country == "Venezuela (Bolivarian Republic of)" ~ "Venezuela",
        country == "Viet Nam"                   ~ "Vietnam",
        TRUE ~ country
      )),
    by = "region",
    relationship = "many-to-many"
  )
```

------------------------------------------------------------------------

### Global cases:

**World map of Tuberculosis Preventive Treatment (TPT):**

TPT = treatment which is given to people who have a dormant TB infection, but who are at risk of getting an active TB infection.

```{r}

# Filter for 2024
map_2024 <- map_TB_10_years_joined |>
  filter(year == 2024)

# Plot
ggplot(map_2024, aes(x = long, y = lat, group = group)) +
  geom_polygon(aes(fill = Preventive_Tx_Pct), color = "white", size = 0.05) +
  
  # Your Custom Red-to-Green Scale
  scale_fill_gradientn(
    colors = c("#A50026", "#D73027", "#F46D43", "#FDAE61", "#FEE08B", 
               "#FFFFBF", "#D9EF8B", "#A6D96A", "#66BD63", "#1A9850", "#006837"),
    na.value = "grey90",
    name = "Coverage (%)",
    limits = c(0, 100),
    labels = label_percent(scale = 1)
  ) +
  
  labs(
    title = "TB Preventive Treatment Coverage (2024)",
    subtitle = "Percentage of estimated contacts provided with TPT"
  ) +
  theme_void() +
  coord_fixed(1.3) +
  theme(
    legend.position = "bottom",
    legend.key.width = unit(2, "cm"),
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )
```

```{r}
# 1. Filter the data (2016 to 2024)
map_filtered <- map_TB_10_years_joined |>
  filter(year >= 2016 & year <= 2024)

# 2. Plot the Grid
ggplot(map_filtered, aes(x = long, y = lat, group = group)) +
  geom_polygon(aes(fill = Preventive_Tx_Pct), color = "white", size = 0.05) +
  
  # Red-to-Green Color Scale
  scale_fill_gradientn(
    colors = c("#A50026", "#D73027", "#F46D43", "#FDAE61", "#FEE08B", 
               "#FFFFBF", "#D9EF8B", "#A6D96A", "#66BD63", "#1A9850", "#006837"),
    na.value = "grey90",
    name = "Coverage (%)",
    limits = c(0, 100),
    labels = label_percent(scale = 1)
  ) +
  
  # Grid Layout
  facet_wrap(~year, ncol = 3) + 
  
  labs(
    title = "Evolution of TB Preventive Treatment Coverage (2016-2024)",
  ) +
  theme_void() +
  coord_fixed(1.3) +
  theme(
    legend.position = "bottom",
    legend.key.width = unit(2.5, "cm"),
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    strip.text = element_text(face = "bold", size = 12)
  )
```

```{r}
# Global analysis (2024)
analysis_2024 <- TB_10_years_joined |>
  filter(year == 2024) |>
  summarise(
    # 1. Estimated Household Contacts
    Global_Estimated_Contacts = sum(Contacts_best, na.rm = TRUE),
    
    # 2. People Provided TPT
    # Calculated as Estimated * Coverage %
    Global_TPT_Total = sum(Contacts_best * (Preventive_Tx_Pct / 100), na.rm = TRUE),
    
    # 3. Children <5 Eligible & on TPT
    Global_Kids_Eligible = sum(Preventive_Tx_Eligible, na.rm = TRUE),
    Global_Kids_TPT = sum(Preventive_Tx_Eligible * (Preventive_Tx_Kids_Pct / 100), na.rm = TRUE)
  ) |>
  mutate(
    Global_Coverage_Pct = (Global_TPT_Total / Global_Estimated_Contacts) * 100
  )

print(analysis_2024)
```

**World map of TB cases pr. 100k citizens:**

```{r}

# Filter for 2024
map_2024 <- map_TB_10_years_joined |>
  filter(year == 2024)

# Plot
ggplot(map_2024, aes(x = long, y = lat, group = group)) +
  geom_polygon(aes(fill = e_inc_100k), color = "white", size = 0.1) +
  # Using 'viridis' option 'inferno' or 'magma' for high contrast
  scale_fill_viridis_c(
    option = "magma", 
    direction = -1, # Darker colors for higher incidence
    name = "Incidence\n(per 100k)",
    na.value = "grey90"
  ) +
  labs(
    title = "Global TB Incidence (per 100K) (2024)",
  ) +
  theme_void() +
  theme(legend.position = "bottom", legend.key.width = unit(2, "cm")) +
  coord_fixed(1.3)

```

```{r}
# 1. Filter the data (2016 to 2024)
map_filtered <- map_TB_10_years_joined |>
  filter(year >= 2016 & year <= 2024)

# 2. Plot the Grid
ggplot(map_filtered, aes(x = long, y = lat, group = group)) +
  geom_polygon(aes(fill = e_inc_100k), color = "white", size = 0.05) +
  
  # Red-to-Green Color Scale
  scale_fill_viridis_c(
    option = "magma", 
    direction = -1, # Darker colors for higher incidence
    name = "Incidence\n(per 100k)",
    na.value = "grey90"
  ) +
  
  # Grid Layout
  facet_wrap(~year, ncol = 3) + 
  
  labs(
    title = "Evolution of TB Incidence (per 100K) (2016-2024)",
  ) +
  theme_void() +
  coord_fixed(1.3) +
  theme(
    legend.position = "bottom",
    legend.key.width = unit(2.5, "cm"),
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    strip.text = element_text(face = "bold", size = 12)
  )
```

**Bar chart - Total amount of people provided with TB Preventive Treatment (TPT):**

```{r}
# 1. Load Data
df <- TB_10_years_joined

# 2. Data Processing
# We need to calculate the absolute numbers from the percentages and denominators.
plot_data <- df |>
  mutate(
    # A. Calculate Number of Children (<5) on TPT
    # 'Preventive_Tx_Eligible' is typically the number of eligible children <5
    # 'Preventive_Tx_Kids_Pct' is the coverage % for this group
    TPT_Kids = Preventive_Tx_Eligible * (Preventive_Tx_Kids_Pct / 100),
    
    # B. Calculate Total Household Contacts on TPT
    # 'Contacts_best' is the number of contacts evaluated (denominator)
    # 'Preventive_Tx_Pct' is the overall contact coverage %
    TPT_Total_Contacts = Contacts_best * (Preventive_Tx_Pct / 100)
  ) |>
  # Aggregate global totals by year
  group_by(year) |>
  summarise(
    TPT_Kids_Global = sum(TPT_Kids, na.rm = TRUE),
    TPT_Total_Global = sum(TPT_Total_Contacts, na.rm = TRUE)
  ) |>
  mutate(
    # C. Derive Adults/Older Contacts (>=5) on TPT
    # Subtract Kids from the Total. Ensure no negative values.
    TPT_Adults_Global = pmax(0, TPT_Total_Global - TPT_Kids_Global)
  ) |>
  # Reshape data for plotting (Long format)
  select(year, TPT_Kids_Global, TPT_Adults_Global) |>
  pivot_longer(
    cols = c("TPT_Kids_Global", "TPT_Adults_Global"),
    names_to = "Category",
    values_to = "Number_on_TPT"
  ) |>
  # Rename categories for the legend
  mutate(
    Category = case_when(
      Category == "TPT_Kids_Global" ~ "Children aged <5 years",
      Category == "TPT_Adults_Global" ~ "Household contacts aged ≥5 years"
    ),
    # Set factor levels to control stacking order (Adults on bottom, Kids on top)
    Category = factor(Category, levels = c("Children aged <5 years", "Household contacts aged ≥5 years"))
  )

# 3. Create the Plot
# Modeled after WHO Global TB Report Fig 3.1 styles
ggplot(plot_data, aes(x = year, y = Number_on_TPT, fill = Category)) +
  geom_col(width = 0.7) +
  # Use colors similar to the WHO report (Orange for Kids, Light Blue for Adults)
  scale_fill_manual(values = c(
    "Children aged <5 years" = "#E69F00",       # Orange
    "Household contacts aged ≥5 years" = "#56B4E9" # Light Blue
  )) +
  scale_y_continuous(
    labels = label_number(scale = 1e-6, suffix = "M"),
    expand = expansion(mult = c(0, 0.1))
  ) +
  scale_x_continuous(breaks = 2015:2024) +
  labs(
    title = "Global Number of People Provided with TB Preventive Treatment",
    subtitle = "Household Contacts",
    x = "Year",
    y = "Number of people (Millions)",
    fill = "Target Group: ",
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid.major.x = element_blank(),
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 11)
  )
```

**Line plot - TB preventive treatment (TPT) coverage, comparing overall household contacts with children under 5 years:**

```{r}
# Load and Prep Data
df <- TB_10_years_joined
df_clean <- df |>
  mutate(TPT_Contacts_Num = Contacts_best * Preventive_Tx_Pct / 100)

global_trends <- df_clean |>
  filter(year >= 2015) |>
  group_by(year) |>
  summarise(
    Total_TPT_Contacts = sum(TPT_Contacts_Num, na.rm = TRUE),
    Total_Contacts_Eligible = sum(Contacts_best, na.rm = TRUE),
    Avg_Child_Coverage = mean(Preventive_Tx_Kids_Pct, na.rm = TRUE)
  ) |>
  mutate(Contact_Coverage = (Total_TPT_Contacts / Total_Contacts_Eligible) * 100)

# Generate Plot
p <- ggplot(global_trends, aes(x = year)) +
  geom_line(aes(y = Contact_Coverage, color = "Household Contacts (All Ages)"), size = 1.2) +
  geom_point(aes(y = Contact_Coverage, color = "Household Contacts (All Ages)"), size = 3) +
  geom_line(aes(y = Avg_Child_Coverage, color = "Children (<5 years)"), linetype = "dashed", size = 1.2) +
  geom_point(aes(y = Avg_Child_Coverage, color = "Children (<5 years)"), shape = 15, size = 3) +
  scale_x_continuous(breaks = 2015:2024) +
  scale_y_continuous(limits = c(0, 100), labels = label_percent(scale = 1)) +
  labs(
    title = "Trends in TB Preventive Treatment Coverage",
    subtitle = "Comparison of Overall Contacts vs. Children (<5y)",
    y = "Coverage (%)",
    x = "Year",
    color = "Target Group: "
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

print(p)
```

Children under 5 years of age are a priority group for TB preventive treatment (TPT), because they are at significantly higher risk of rapid progression from TB infection to active disease and often develop more severe forms of TB.

Therefore, this helps explain why we see more kids having received TPT.

```{r}
# 1. Prepare the Data
gap_analysis <- TB_10_years_joined |>
  filter(year == 2024) |>
  filter(!is.na(Preventive_Tx_Pct) & !is.na(Contacts_best))

# 2. Define Dynamic Thresholds
# Burden Threshold: Top 10% of countries with the most contacts
burden_threshold <- quantile(gap_analysis$Contacts_best, 0.90) 

# Coverage Threshold: Less than 40% coverage (Performance Gap)
coverage_threshold <- 40

# 3. Create a Flag for High Risk Countries
gap_analysis <- gap_analysis |>
  mutate(
    Is_High_Risk = if_else(Contacts_best >= burden_threshold 
                           & Preventive_Tx_Pct < coverage_threshold, 
                           "High Risk (High Burden, Low Coverage)", 
                           "Other")
  )

# 4. Plot
ggplot(gap_analysis, aes(x = Contacts_best, y = Preventive_Tx_Pct)) +
  # Plot "Other" points first (grey)
  geom_point(data = subset(gap_analysis, Is_High_Risk == "Other"), 
             alpha = 0.5, color = "darkgrey", size = 2) +
  
  # Plot "High Risk" points on top (Dynamic Color)
  geom_point(data = subset(gap_analysis, Is_High_Risk != "Other"), 
             aes(color = Is_High_Risk), size = 3) +
  
  # Add labels ONLY for the High Risk countries
  geom_text(data = subset(gap_analysis, Is_High_Risk != "Other"), 
            aes(label = country), vjust = -1, size = 3, fontface = "bold") +

  # Formatting
  scale_x_log10(labels = scales::comma) + 
  scale_y_continuous(limits = c(0, 100)) +
  scale_color_manual(values = c("High Risk (High Burden, Low Coverage)" = "firebrick")) +
  
  labs(
    title = "The Gap: Burden vs. Treatment Performance",
    subtitle = paste0("Highlighted: Top 10% Burden (> ", comma(burden_threshold), " contacts) AND Coverage < 40%"),
    x = "Household Contacts (Log Scale)",
    y = "Preventive Treatment Coverage (%)",
    color = "Country Status"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

This gap plot helps us identify countries where there are a high TB household contact burden, but where the TB preventive treatment (TPT) is low (insufficient coverage).

I.e. these countries are faced with the problem: the burden is high, but treatment coverage is low.

### **AMR resistance insights:**

One of the first interesting questions that arise is how is the trending of AMR in TB.

```{r}
TB_10_years_joined |>
  group_by(year) |>
  summarise(
    New = mean(rr_new, na.rm = TRUE),
    Treated = mean(rr_treated, na.rm = TRUE),
    Incidence = mean(rr_incidence, na.rm = TRUE)) |>
  pivot_longer(cols = c(New, Treated, Incidence), 
               names_to = 'type', values_to = 'value') |>
  
  ggplot(aes(x = year, y = value)) +
  geom_point(alpha = 0.45) +
  geom_smooth(aes(color = type)) +
  scale_color_brewer(palette = "Set2") +
  facet_wrap(
    ~ type, 
    ncol = 1, 
    scales = 'free_y',
    strip.position = "left",
    labeller = labeller(type = c(
      "New" = "Percentage (%)",
      "Treated" = "Percentage (%)",
      "Incidence" = "Amount of Cases"
    ))
  ) +
  scale_x_continuous(
    breaks = unique(TB_10_years_joined$year)  #Ensures that the x-axis is shown as integer years. As we don't want it as decimal numbers. 
  ) +
  labs(
    x = "Year",
    y = "",
    title = "Amount of new(%), treated(%) and TB incidences cases\n Global timeline",
    color = "Type"
  ) +
  theme_minimal()

```

With this previous plot the results are expected and promising, even though the amount is minimal there's a diminution of cases each year.

It could look like the "New" incidents its starting to go up again - but more information would be needed for that, as its only year 2024 that had an increase.

Another question of interest its if there's any relationship between the drug resistance and the mortality, and furthermore, if its something global or its per region.

**Drug resistance relative to mortality:**

```{r}
r2_by_region <- TB_10_years_joined |>
  group_by(g_whoregion.x) |>
  summarise(
    r2 = summary(lm(e_mort_100k ~ I(rr_incidence/e_pop_num*1e5)))$r.squared
  )
r2_by_region
```

```{r}
labels <- r2_by_region |>
  mutate(
    label = str_c(g_whoregion.x, " (R² = ", round(r2, 2), ")")
  )
```

```{r}
TB_10_years_joined |>
  ggplot(aes(x=(rr_incidence/e_pop_num*10^5),y=e_mort_100k, color=g_whoregion.x)) +
  geom_point(alpha = 0.45)+
  geom_smooth(method='lm', se = FALSE)+
  coord_cartesian(xlim=c(0,45), ylim=c(0,150)) +  #Eye calculated good limit for the plot
  scale_color_brewer(palette = "Set2",
    labels = labels$label,
    breaks = labels$g_whoregion.x) +
  labs(
        x = "RR - TB", 
        y = "Mortality", 
        title = "Relation between RR - TB and Mortality per region", 
        color = "WHO Region") +
  theme_minimal()
```

AFR = Africa

AMR = America

EMR = Eastern Mediterranean Region

EUR = Europe

SEA = South-East Asia

WPR = Western Pacific Region

This graph shows a clear linear tendency between mortality to increase relative to the increase of drug resistance.

It's also interesting how the regions that control better drug resistance, as Europe, have less mortality, which of course is an expected outcome.

Meanwhile, Eastern Mediterranean Region worst control of the mortality relative to drug resistance.

### **Household insights:**

```{r}
# Filter for 2024
map_2024 <- map_TB_10_years_joined |>
  filter(year == 2024)

ggplot(map_2024, aes(long, lat, group = group))+
  geom_polygon(aes(fill = Contacts_best ), color = "white", size = 0.1) +
  coord_quickmap() + 
  scale_fill_viridis_c(option = "C") +
  labs(
        x = "", 
        y = "", 
        title = "Household TB contacts by country (2024)", 
        fill = "Contacts") +
  theme_void() +
  theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5),
      legend.position = "bottom",
  legend.key.width = unit(2, "cm"))
```

The map shows how India is the country with most TB contacts during 2024, which is perhaps not too surprising, given that it is an overcrowded country.

```{r}
# 1. Filter data for the grid (2016-2024)
map_contacts_grid <- map_TB_10_years_joined |>
  filter(year >= 2016 & year <= 2024)

# 2. Plot the Grid
ggplot(map_contacts_grid, aes(long, lat, group = group)) +
  
  geom_polygon(aes(fill = Contacts_best), color = "white", size = 0.05) +
  
  coord_quickmap() +
  
  scale_fill_viridis_c(option = "C") +
  
  # Create the Grid
  facet_wrap(~year, ncol = 3) +
  
  labs(
    title = "Household TB Contacts by Country (2016-2024)",
    subtitle = "Logarithmic Scale: Brighter colors indicate higher number of contacts",
    x = "", 
    y = ""
  ) +
  
  theme_void() +
  
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    legend.position = "bottom",
    legend.key.width = unit(2.5, "cm"),
    strip.text = element_text(face = "bold", size = 12) # Makes the year labels readable
  )
```

Another hypothesis to prove was that if the amount of detection could contribute to lower the amount of household contacts:

**Detection rate relative to household contacts pr. 100k citizens:**

```{r}
TB_10_years_joined |>
  ggplot(aes(x = (Contacts_best / e_pop_num) * 10^5, y = c_cdr_pct)) +
  geom_point(alpha = 0.25) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(
    x = "Household contacts per 100k",
    y = "Case detection rate (%)",
    subtitle = "Relationship between household contacts and CDR"
  ) +
  theme_minimal()
```

The plot shows that the hypothesis was correct; countries with high detection rates (%) have a tendency to have lower household contacts pr. 100k.

It is also interesting to there are some outlayers in the scatter plot, given that the maximum should be 100% but for some reason there are values as high as \~140%.

### **HIV insights:**

It's easy to conclude that having TB, HIV and RR cause more fatalities than having only TB. But comparing those who have TB and HIV from those who have RR - TB is an interesting topic:

```{r}
TB_10_years_joined |>
mutate(
disease = case_when(
  e_tbhiv_prct > mean(e_tbhiv_prct, na.rm = TRUE) & rr_new > mean(rr_new, na.rm = TRUE) ~ "HIV & RR-TB",
  e_tbhiv_prct > mean(e_tbhiv_prct, na.rm = TRUE) ~ "HIV & TB",
  rr_new > mean(rr_new, na.rm = TRUE) ~ "Only RR-TB",
  TRUE ~ "Only TB"
)) |>
ggplot(aes(x=disease, y =e_mort_100k, fill = disease))+
geom_boxplot() +
scale_fill_brewer(palette = "Set2") + 
labs(
    x = "",
    y = "Mortality cases per 100k",
    subtitle = "Mortality according to combinations of HIV and RR-TB",
    fill = "Disease"
  ) +
  theme_minimal()
```

From this boxplot shows that there might be a weak tendency of HIV and RR-TB to cause more fatalities, relative to individuals who "only" have TB.

Nevertheless is interesting to find that there's more mortality in cases of HIV and TB than in cases of RR-TB.

More data is needed to conclude the exact reason, but the most probable reason is the synergy between TB and HIV. Having HIV makes the immune system struggle to fight TB, and then TB can cause more inflamation which at the same time makes easier for HIV to replicate creating a negative feedback loop.
